The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-10-22 theoretical-consensus/article/article.tex
add page 04

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
...
\end{abstract}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The first asynchronous consensus algorithm was later proposed by Castro and Liskov in \cite{castro1999practical}. 
And naturally, after that work, many other asynchronous consensus algorithms have appeared. 
However, all of them are based on the assumption that the number of faulty processes is less than 
a certain threshold. Additionally, the assumption of a known set of participants is also made, as well as their roles 
in the consensus protocol. These are very strong assumptions that limit the challenges that can be addressed, 
for example, the impossibility to know the participants beforehand as they may participate anonymously, or dynamically.

@@

Following the line of the previous sections, it has been reasoned about the need 
for permissionless consensus when there are already well known and established consensus protocols
that work in trusted environments \cite{castro1999practical, miller2016honey}. However, even those protocols have
their own limitations, not only in terms of trust, fault-tolerance, centrality, permissions,
or bottlenecks, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
in size. The more participants there are, the more messages need to be exchanged
between them, and the more time it takes to reach consensus, even if assuring deterministic finality \cite{decker2016bitcoin}. 
This is a problem that is not only related to the number of participants, but also to the
communication fashion and bandwidth, and the computational capacity of the devices that
participate in the consensus protocol. 
Summing up, the need for permissionless consensus is then justified by the fact that
permissioned protocols are not compatible with the requirements of the new generation of
distributed systems, especially in the context of blockchain networks. 
These requirements include dealing with today's sparse networks of anonymously and dynamically
participating devices, without interrupting consensus and while battling Sybil attacks \cite{survey-dist-consensus}.
Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
and where the network is sparse and unpredictably unreliable.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by chaining the generation of the agreeable value, 
in this particular case, a block and its proof, with the dissemination of 
the information to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
for further network agreement, complies with the unit of computational work needed to
create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
This essentially requires brute forcing the search for a cryptographic hash value for the
aggregation of the block information with a nonce, such that this hash value satisfies
a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
requirement for the block generation interval \cite{survey-dist-consensus}.

There are a couple of particularities that need to be mentioned. First, the block generation
is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
also plays a role in this regard, by naturally incentivizing honest participation.
The block generation interval, forged by the adjustment of the difficulty
threshold, is also of major importance because it reduces the fork probability and
allows for the timely propagation of proposed blocks, guaranteeing consensus
in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.

@@

Despite the harmony of all the aforementioned clever characteristics,
the proof-of-work mechanism has some well known drawbacks. 
First, the block generation is computationally expensive, and so the
energy consumption associated with it is also high. This is caused by the
dynamically adjusted difficulty threshold, which is a function of the block 
generation interval. For example, in the Bitcoin network, to maintain the 10 
minutes interval, the hashing difficulty
increases with the increase of the hashing power, and so too the energy needed for it. 
However, reducing the block generation interval would increase the
probability of forks, which results in a more frequent waste of useful work,
loosening the security of the network. Increasing the block size would also
have the same security effect due to the increased propagation delay.

@@

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submited yet to the Distributed Systems Seminar. The content of this paper is subject to change.
\end{abstract}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort, such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The first practical asynchronous consensus algorithm was later proposed by Castro and Liskov in \cite{castro1999practical}. 
And naturally, after that work, many other asynchronous consensus algorithms have appeared. 
However, all of them are based on the assumption that the number of faulty processes is less than 
a certain threshold. Additionally, the assumption of a known set of participants is also made, as well as their roles 
in the consensus protocol. These are very strong assumptions that limit the challenges that can be addressed, 
for example, the impossibility to know the participants beforehand as they may participate anonymously, or dynamically.

@@

Following the line of the previous sections, it has been reasoned about the need 
for permissionless consensus when there are already well known and established consensus protocols
that work in trusted environments \cite{castro1999practical, miller2016honey}. However, even those protocols have
their own limitations, not only in terms of trust, fault-tolerance, centrality, permissions,
or bottlenecks, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
in size. The more participants there are, the more messages need to be exchanged
between them, and the more time it takes to reach consensus, even if assuring deterministic finality \cite{decker2016bitcoin}. 
This is a problem that is not only related to the number of participants, but also to the
communication fashion and bandwidth, and the computational capacity of the devices that
participate in the consensus protocol. 
Summing up, the need for permissionless consensus is then justified by the fact that
permissioned protocols are not compatible with the requirements of the new generation of
distributed systems, especially in the context of blockchain networks. 
These requirements include dealing with today's sparse networks of anonymously and dynamically
participating devices, without interrupting consensus and while battling Sybil attacks \cite{survey-dist-consensus}.
Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
and where the network is bigger, sparse and unpredictably unreliable.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by first generating the agreeable value, 
in this particular case, a block and its proof, disseminating the information 
to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

@@

In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
for further network agreement, complies with the unit of computational work needed to
create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
This essentially requires brute forcing the search for a cryptographic hash value for the
aggregation of the block information with a nonce, such that this hash value satisfies
a difficulty threshold (See Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
requirement for the block generation interval \cite{survey-dist-consensus}.

There are a couple of particularities that need to be mentioned. First, the block generation
is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
Allied to this, increasing the hashing power is not an easy task, which consequently allows for tackling Sybil attacks,
despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
also plays a role in this regard, by naturally incentivizing honest participation.
The block generation interval, forged by the adjustment of the difficulty
threshold, is also of major importance because it reduces the fork probability and
allows for the timely propagation of proposed blocks, guaranteeing consensus
in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.

@@

Despite the harmony of all the aforementioned clever characteristics,
the proof-of-work mechanism has some well known drawbacks. 
First, the block generation is computationally expensive, and so, the
energy consumption associated with it is also high. This is caused by the
dynamically adjusted difficulty threshold, which is a function of the block 
generation interval. For example, in the Bitcoin network, to maintain the 10 
minutes interval, the hashing difficulty
increases with the increase of the hashing power, and so too the energy needed for it. 
However, reducing the block generation interval would increase the
probability of forks, which results in a more frequent waste of useful work,
loosening the security of the network. Increasing the block size would also
have the same security effect due to the increased propagation delay.

@@

\section{Alternatives to Proof-of-Work}

\subsection{Proof-of-Stake based protocols}

At this point, we can exercise the imagination and extrapolate the
previous block generation mechanism to a \emph{Proof-of-Something} 
pseudo-random competition in which an entity in possession of a higher
amount of a certain resource, either computational power, or stake, 
or certain currency, or a higher amount of storage space, etc., guarantees
a higher probability of leading the block generation and proposal, and
consequently winning acceptance by the majority. This is the essence of
the Proof-of-Stake consensus protocol, as a derivative of the Proof-of-Work
mechanism, where a stake is a traceable and verifiable amount of a certain
unit, token or currency, that is owned by a certain entity that wishes to
participate in the consensus protocol. The stake works as a form of collateral
that is used to guarantee the honesty of the entities, in an attempt to
reduce the Sybil attack probability. And, generally as in Proof-of-Work, 
the higher the stake, the higher the probability of leading the block generation 
and proposal. Proof-of-Stake is, indeed, a more energy efficient consensus protocol,
however, this adds up to nothing, especially when compared to the already proved
BFT solutions, if centralization is a functional requirement, 
or unfair advantage is a potential trait. 
These limitations, along with other security concerns, may push away the 
participation of resourceless entities, and consequently, the network
decentralization.

\subsubsection{Chain-based and Committee-based protocols}

Due to the self-contained nature of this type of protocols, where the
attestation of the stake is done inside the network, multiple classes
of Proof-of-Stake can be distinguished, by the way the meaningfulness of
the stake is defined and employed. For example, there are the simplest forms of \emph{PoS} that inherit most
of the characteristics of the Proof-of-Work mechanism, but with the
stake as part of the block generation and hashing proof (See Algorithm \ref{alg:PoSBlockGeneration}).
These are commonly referred to as Chain-based \emph{PoS}, inheriting also the
fault-tolerance properties of the Nakamoto consensus protocol.
Committee-based \emph{PoS} protocols, on the other hand, are more complex,
and they are based on the idea of a committee of entities that are 
selected to lead the block generation and proposal. The committee is
formed by a subset of the network entities, and the selection is done
based on their stake. Usually, the selection takes a form of a random
sequence of stake holders, where the probability of being selected
for every spot is proportional to the stake. That sequence is then 
consciously timed to orderly pace the block generation and proposal, 
by the current slot leader. Examples of this type of mechanisms are
the \emph{Ouroboros} \cite{kiayias2017ouroboros}, \emph{Ouroboros Praos} \cite{david2018ouroboros} protocols
or \emph{Snow White} \cite{daian2019snow}. The longest chain rule is
still applied, with probabilistic finality, and the 1/2 threshold is
still required or tolerated for the network to reach consensus. Nonetheless,
Committee-based \emph{PoS} protocols may not pair well with decentralization and 
increase in the committee size, which detereorates the performance of the
protocol. Several improvements have been proposed to tackle not only this
issue, but also the security of the committees which, smaller, could be more
vulnerable to targeted attacks \cite{david2018ouroboros}.

\begin{algorithm}
  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
  \begin{algorithmic}[1]
    \Function {}{}
      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
      \Indent
        \State $| \ Hash \ of \ the \ last \ Block$
        \State $| \ Timestamp$
        \State $| \ Other$
      \EndIndent

      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
      \State Increment $clockTime$ by a constant $tickInterval$
      \EndWhile
      \State \Return new block
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\subsubsection{BFT-based protocols}

These aim at achieving deterministic finality, by employing a BFT
consensus protocol, not at the block proposal level, but at the
block acceptance level. The idea is to have a free flow of blocks
proposed by the network, but to have a quorum of entities that
are responsible for the block acceptance. The block proposal mechanism
can be of any kind, for example, Chain-based \emph{PoS}, or Committee-based \emph{PoS},
or even pure Proof-of-Work. The block acceptance layer then allows
for a deterministic acceptance of the blocks, by having a sort of permissioned
BFT consensus protocol, with the trivial 1/3 byzantine tolerance. The most well known examples
and variations are the Tendermint \cite{buchman2016tendermint} and the new Ethereum Casper 
\cite{buterin2017casper} protocols.

Delegated Proof-of-Stake (DPoS) is a variation of the BFT-based protocols,
where the consensus group is formed by a subset of the network entities,
which are selected by the network, with a voting mechanism based on the
delegation of stake. The consensus is then achieved using a typical 
PBFT protocol, with the usual 1/3 byzantine tolerance.

\subsection{}
